---
title: "Data Wrangling with dplyr and tidyr"

teaching: 20
exercises: 10

---

:::: questions

- How can I select specific rows and/or columns from a data-frame?
- How can I combine multiple commands into a single command?
- How can I create new columns or remove existing columns from a dataframe?
- How can I reformat a dataframe to meet my needs?

::::


:::: objectives

- Describe the purpose of an R package and the **`dplyr`** and **`tidyr`** packages.
- Select certain columns in a dataframe with the **`dplyr`** function `select`.
- Select certain rows in a dataframe according to filtering conditions with the **`dplyr`**
  function `filter`.
- Link the output of one **`dplyr`** function to the input of another function with
  the 'pipe' operator ` |> `.
- Add new columns to a dataframe that are functions of existing columns with `mutate`.
- Use the split-apply-combine concept for data analysis.
- Use `summarize`, `group_by`, and `count` to split a dataframe into groups of observations,
  apply a summary statistics for each group, and then combine the results.
- Describe the concept of a wide and a long table format and for which purpose those
  formats are useful.
- Describe the roles of variable names and their associated values when a table is
  reshaped.
- Reshape a dataframe from long to wide format and back with the `pivot_wider` and
  `pivot_longer` commands from the **`tidyr`** package.
- Export a dataframe to a csv file.

::::

```{r, include = FALSE}
library(dplyr)
library(readr)
movieSerie <- read_csv("data/movieSerie.csv", na = c("NA", "NULL"))
#interviews <- read_csv("../data/SAFI_clean.csv", na = "NULL")
```

**`dplyr`** is a package for making tabular data wrangling easier by using a
limited set of functions that can be combined to extract and summarize insights
from your data. It pairs nicely with **`tidyr`** which enables you to swiftly
convert between different data formats (long vs. wide) for plotting and analysis.

Similarly to **`readr`**, **`dplyr`** and **`tidyr`** are also part of the
tidyverse. These packages were loaded in R's memory when we called
`library(tidyverse)` earlier.

:::: callout

## Note

 The packages in the tidyverse, namely **`dplyr`**, **`tidyr`** and **`ggplot2`**
 accept both the British (e.g. *summarise*) and American (e.g. *summarize*) spelling
 variants of different function and option names. For this lesson, we utilize
 the American spellings of different functions; however, feel free to use
 the regional variant for where you are teaching.

::::::

## What is an R package?

The package **`dplyr`** provides easy tools for the most common data
wrangling tasks. It is built to work directly with dataframes, with many
common tasks optimized by being written in a compiled language (C++) (not all R
packages are written in R!).

The package **`tidyr`** addresses the common problem of wanting to reshape your
data for plotting and use by different R functions. Sometimes we want data sets
where we have one row per measurement. Sometimes we want a dataframe where each
measurement type has its own column, and rows are instead more aggregated
groups. Moving back and forth between these formats is nontrivial, and
**`tidyr`** gives you tools for this and more sophisticated data wrangling.

But there are also packages available for a wide range of tasks including
building plots (**`ggplot2`**, which we'll see later), downloading data from the
NCBI database, or performing statistical analysis on your data set. Many
packages such as these are housed on, and downloadable from, the
**C**omprehensive **R** **A**rchive **N**etwork (CRAN) using `install.packages`.
This function makes the package accessible by your R installation with the
command `library()`, as you did with `tidyverse` earlier.

To easily access the documentation for a package within R or RStudio, use
`help(package = "package_name")`.

To learn more about **`dplyr`** and **`tidyr`** after the workshop, you may want
to check out this [handy data transformation with **`dplyr`** cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
and this [one about **`tidyr`**](https://rstudio.github.io/cheatsheets/html/tidyr.html).

## Learning **`dplyr`** and **`tidyr`**

We are working with the same dataset as earlier. Refer to the 
previous lesson to download the data, if you do not have it loaded.

We're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows

To select columns of a dataframe, use `select()`. The first argument to this
function is the dataframe (`movieSerie`), and the subsequent arguments are the
columns to keep, separated by commas. Alternatively, if you are selecting
columns adjacent to each other, you can use a `:` to select a range of columns,
read as "select columns from ___ to ___."

```{r, purl = FALSE}
# to select columns throughout the dataframe
select(movieSerie, title, description)
# to select a series of connected columns
select(movieSerie, title:description)
```

To choose rows based on specific criteria, we can use the `filter()` function.
The argument after the dataframe is the condition we want our final
dataframe to adhere to (e.g. age_certification is PG-13): 

```{r, purl = FALSE}
# filters observations where age_certification name is "PG-13" 
filter(movieSerie, age_certification == "PG-13")
```

We can also specify multiple conditions within the `filter()` function. We can
combine conditions using either "and" or "or" statements. In an "and" 
statement, an observation (row) must meet **every** criteria to be included
in the resulting dataframe. To form "and" statements within dplyr, we can  pass
our desired conditions as arguments in the `filter()` function, separated by
commas:

```{r, purl=FALSE}

# filters observations with "and" operator (comma)
# output dataframe satisfies ALL specified conditions
filter(movieSerie, age_certification == "PG-13",
                   runtime > 100,
                   imdb_score < 6.0)
```

We can also form "and" statements with the `&` operator instead of commas:

```{r, purl=FALSE}
# filters observations with "&" logical operator
# output dataframe satisfies ALL specified conditions
filter(movieSerie, age_certification == "PG-13" & 
                   runtime > 100 & 
                   imdb_score < 6.0)
```

In an "or" statement, observations must meet *at least one* of the specified conditions. 
To form "or" statements we use the logical operator for "or," which is the vertical bar (|): 

```{r, purl=FALSE}
# filters observations with "|" logical operator
# output dataframe satisfies AT LEAST ONE of the specified conditions
filter(movieSerie, age_certification == "PG-13" | 
                   runtime > 100 | 
                   imdb_score < 6.0)
```


## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary dataframe and use
that as input to the next function, like this:

```{r, purl = FALSE}
movieSerie2 <- filter(movieSerie, age_certification == "PG-13")
movieSerie_ch <- select(movieSerie2, title:description)
```

This is readable, but can clutter up your workspace with lots of objects that
you have to name individually. With multiple steps, that can be hard to keep
track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl = FALSE}
movieSerie_ch <- select(filter(movieSerie, age_certification == "PG-13"),
                         title:description)
```

This is handy, as R evaluates the expression from the inside out (in this case, 
filtering, then selecting), but it can be difficult to read if too many 
functions are nested, 

The last option is *pipes*. Pipes let you take the
output of one function and send it directly to the next, which is useful when
you need to do many things to the same dataset. The pipe is build into R and looks like ` |> `. If you use Positron, you can type the pipe with:  
- <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> +
<kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac. 

A previous incarnation looks like `%>%`. Both pipes work the exact same way.

If you are using RStudio, the above keyboard shortcuts will result in `%>%`. 

```{r, purl = FALSE}
movieSerie |> 
    filter(age_certification == "PG-13") |> 
    select(title,description)
```

In the above code, we use the pipe to send the `movieSerie` dataset first
through `filter()` to keep rows where `age_certification` is "PG-13", then through
`select()` to keep only the `title` and `description` columns. Since ` |> `
takes the object on its left and passes it as the first argument to the function
on its right, we don't need to explicitly include the dataframe as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we take the dataframe `movieSerie`, *then* we `filter`
for rows with `age_certification == "PG-13"`, *then* we `select` columns `title` and
`description`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex data wrangling operations.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl = FALSE}
movieSerie_ch <- movieSerie |> 
    filter(age_certification == "PG-13") |> 
    select(title,description)

movieSerie_ch

```

Note that the final dataframe (`movieSerie_ch`) is the leftmost part of this
expression.


:::: challenge


## Exercise

Using pipes, subset the `movieSerie` data set to include movieSerie
have a `release_year` greater than 1980 and retain only the columns `title`,
 `runtime`, and `age_certification`.
 
::::

:::: solution

## Solution

```{r}
movieSerie |> 
     filter(release_year > 1980) |> 
     select(title, runtime, age_certification)
```

::::


## Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in
two columns. For this we'll use `mutate()`.

We might be interested in knowing the differences in scores on imdb vs tmdb:

```{r, purl = FALSE}
movieSerie |> 
  mutate(score_difference = imdb_score - tmdb_score) |> 
  select(imdb_score, tmdb_score, score_difference)
```


:::: challenge 

## Exercise

Create a new dataframe from the `movieSerie` data set that meets the following
criteria: contains only the `title` column and a new column called
`total_score` containing a value that is equal to the total number of scores on 
both imdb and tmdb (`imdb_score` plus `tmdb_score`).
Only the rows where `total_score` is greater than 15 should be shown in the
final dataframe.

**Hint**: think about how the commands should be ordered to produce the data
frame!

:::: solution

## Solution

``` {r}
movieSerie_total_score <- movieSerie |> 
  mutate(total_score = imdb_score + tmdb_score) |> 
  filter(total_score > 15) |> 
  select(title, total_score)
```

::::::::
::::

## Split-apply-combine data analysis and the summarize() function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of
the `group_by()` function.


#### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group. `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the average imdb_score by
genre:

```{r, purl = FALSE}
movieSerie |> 
    group_by(genre) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE))
```

You may also have noticed that the output from these calls doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` over dataframe.

You can also group by multiple columns:

```{r, purl = FALSE}
movieSerie |> 
    group_by(genre, type) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE))
```

Note that the output is a grouped tibble. To obtain an ungrouped tibble, use the
`ungroup` function:

```{r, purl = FALSE}
movieSerie |> 
    group_by(genre, type) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE)) |> 
    ungroup()
```



Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the maximum imdb_score given to a movie or serie:

```{r, purl = FALSE}
movieSerie |> 
    group_by(genre, type) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE),
              max_imdb_score = max(imdb_score, na.rm = TRUE))
```

It is sometimes useful to rearrange the result of a query to inspect the values.
For instance, we can sort on `min_imdb_score` to put the group with the smallest
imdb_score first:


```{r, purl = FALSE}
movieSerie |> 
    group_by(genre, type) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE),
              max_imdb_score = max(imdb_score, na.rm = TRUE)) |> 
    arrange(max_imdb_score)
```

To sort in descending order, we need to add the `desc()` function. If we want to
sort the results by decreasing order of minimum imdb_score:

```{r, purl = FALSE}
movieSerie |> 
    group_by(genre, type) |> 
    summarize(mean_imdb_score = mean(imdb_score, na.rm = TRUE),
              max_imdb_score = max(imdb_score, na.rm = TRUE)) |> 
    arrange(desc(max_imdb_score))
```

#### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each village, we would do:

```{r, purl = FALSE}
movieSerie |> 
    count(release_year)
```

For convenience, `count()` provides the `sort` argument to get results in
decreasing order:

```{r, purl = FALSE}
movieSerie |> 
    count(release_year, sort = TRUE)
```


::::::::::::::::::::::::::::::::::::: challenge

## Exercise

1. How many moview and series are there for each age_certification?

2. Use `group_by()` and `summarize()` to find the mean, min, and max for 
tmdb_score. Also add the number of observations (hint: see `?n`).

:::::::::::::::::::::::: solution

## Solution 1

```{r}
 movieSerie |> 
  count(age_certification)
```

:::::::::::::::::::::::::::::::::

:::::::::::::::::::::::: solution

## Solution 2

```{r}
movieSerie |> 
  group_by(genre) |> 
   summarize(
       mean_tmdb_score = mean(tmdb_score, na.rm = TRUE),
       min_tmdb_score = min(tmdb_score, na.rm = TRUE),
       max_tmdb_score = max(tmdb_score, na.rm = TRUE),
       n = n()
   )
```

:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from dataframes.

Before using `write_csv()`, we are going to create a new folder, `data_output`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data_output`
directory, so even if the files it contains are deleted, we can always
re-generate them.

Now we can save this dataframe to our `data_output` directory.

```{r, purl = FALSE, eval = FALSE}
write_csv(movieSerie_ch, file = "data_output/movieSerie_changed.csv")
```

```{r, purl = FALSE, eval = TRUE, echo = FALSE}
if (!dir.exists("../data_output")) dir.create("../data_output")
write_csv(movieSerie_ch, "../data_output/movieSerie_changed.csv")
```

:::: keypoints

- Use the `dplyr` package to manipulate dataframes.
- Use `select()` to choose variables from a dataframe.
- Use `filter()` to choose data based on values.
- Use `group_by()` and `summarize()` to work with subsets of data.
- Use `mutate()` to create new variables.
- Use the `tidyr` package to change the layout of dataframes.
- Use `pivot_wider()` to go from long to wide format.
- Use `pivot_longer()` to go from wide to long format.

::::
